---
title: "R Notebook"
output: html_notebook
---


# Importing the data

Read the raw counts using code from previous session

```{r}
library(tximport)
library(DESeq2)
dirs <- list.files(file.path("salmon_quant/"))
quant_files <- paste0("salmon_quant/",dirs,"/quant.sf.gz")
names(quant_files) <- dirs

tx2gene <- read.csv("tx2gene.csv")

txi <- tximport(quant_files,type="salmon",tx2gene = tx2gene,ignoreTxVersion = TRUE)

sampleinfo <- read.delim("meta_data/sampleInfo_corrected.txt")
rownames(sampleinfo) <- sampleinfo$run

dds <- DESeqDataSetFromTximport(txi, 
                                colData = sampleinfo,
                                design = ~CellType)

```

Analyse the default factor of `CellType` to compare `basal` and `luminal` cell types.

```{r}
de_status <- DESeq(dds)
```

We can use `dplyr` to transform into a more manageable data frame

```{r}
library(dplyr)
library(tibble)
results_cellType <- results(de_status) %>%
  as.data.frame %>% 
  rownames_to_column("ENSEMBL")  

results_cellType
```


# Exercise

We now want to use the `Status` as our factor of interest. This can be achieved by changing the `design` our object

```{r}
dds_status <- dds
colData(dds)
dds_status$Status <- as.factor(dds_status$Status)
design(dds_status) <- ~Status
de_status <- DESeq(dds_status)
```

The default contrast is already virgin vs lactation. As above we can create a data frame for further processing.

```{r}
results_status <- results(de_status) %>%
  as.data.frame %>% 
  rownames_to_column("ENSEMBL")  

results_status
saveRDS(results_status,"Robjects/results_status.rds")
```

We can filter to discover differentially-expressed genes and write out to a file

```{r}
library(readr)
results_status %>% filter(abs(log2FoldChange) > 1, padj < 0.05) %>% 
  write_csv("virgin_vs_lactation_DE.csv")
```

# Exercise

For the venn diagram we will need the results of the two contrasts as data frame. Both these tables can be obtained from the `results_status` object that we already created.

```{r}
library(limma)
results_p_vs_v <- results(de_status, contrast = c("Status","pregnancy","virgin")) %>% 
    as.data.frame %>% 
  rownames_to_column("ENSEMBL")  

results_l_vs_v <- results(de_status, contrast = c("Status","lactation","virgin")) %>% 
    as.data.frame %>% 
  rownames_to_column("ENSEMBL")  

venn_data <- data.frame("Pregnancy" = results_p_vs_v$padj < 0.05, "Lactation" = results_l_vs_v$padj < 0.05)

vennDiagram(venn_data)
```

# Exercise

Performing differential expression on subsets, as-per the notes.

```{r}
## create subset corresponding to just basal samples
dds_basal <- dds[,which(dds$CellType == "basal")]
dds_basal$Status <- as.factor(dds_basal$Status)
design(dds_basal) <- ~Status
## perform analysis
de_basal <- DESeq(dds_basal)
```

```{r}
## create subset corresponding to just luminal samples
dds_luminal <- dds[,which(dds$CellType == "luminal")]
dds_luminal$Status <- as.factor(dds_luminal$Status)
design(dds_luminal) <- ~Status
de_luminal <- DESeq(dds_luminal)
```

The venn diagram is then constructed by making a table of 0 and 1's (depending on whether each gene is differentially-expressed in each analysis).

```{r}
venn_data <- data.frame(Basal = results(de_basal)$padj < 0.05, Luminal  = results(de_luminal)$padj<0.05)

vennDiagram(venn_data)
```
The multi-factor uses all samples, but includes a factor for the `CellType`. We can still extract results for virgin vs lactation, but these are corrected for differences in `CellType`.

```{r}
dds_mf <- dds
dds_mf$Status <- as.factor(dds_mf$Status)
design(dds_mf) <- ~CellType+Status
de_mf <- DESeq(dds_mf)
results_mf <- results(de_mf,contrast=c("Status","virgin","lactation"
                                       ))
results_mf
```

Using a filter we can identify the genes from this analysis. We can see that many more genes identified, so in this case it is better to include all samples and model the difference in `CellType`.

```{r}
filter(data.frame(results_mf), padj < 0.05)
```


