---
title: "R Notebook"
output: html_notebook
---

We start by loading our results from the previous week

```{r}
library(DESeq2)
library(dplyr)
## Read the counts that we produced previously
dds <- readRDS("Robjects/dds.rds")
results_cellType <- readRDS("Robjects/results_cellType.rds")
```

To ease our interpretation we can annotate these using pre-built databases

```{r}
library(org.Mm.eg.db)

  anno <- AnnotationDbi::select(org.Mm.eg.db,keys=results_cellType$ENSEMBL,
              columns=c("ENSEMBL","SYMBOL","GENENAME","ENTREZID"),
              keytype="ENSEMBL") %>% 
  filter(!duplicated(ENSEMBL))
results_annotated <- left_join(results_cellType, anno,by="ENSEMBL") %>% 
  arrange(padj)
```


# Exercise

The `plotCounts` function of `DESeq2` provides a quick diagnostic, but is not particularly attractive. Fortunately it allows us to return the data used in the plot in a format suitable for `ggplot2`

```{r}
library(ggplot2)
library(dplyr)
dds <- readRDS("Robjects/dds.rds")

plotCounts(dds, "ENSMUSG00000027750",intgroup = "CellType",returnData = TRUE) %>% 
  ggplot(aes(x = CellType, y = count,col=CellType)) + geom_jitter(width=0.1) + scale_y_log10()

```

# Exercise

Producing a heatmap of the top 100 genes in the comparison of virgin vs lactation.

```{r}
library(pheatmap)
results_status <- readRDS("Robjects/results_status.rds")
results_annotated <- left_join(results_status, anno,by="ENSEMBL") %>% 
  arrange(padj)

## Set the number of genes as a variable so we can easily change the heatmap

N <- 100
top_genes <- results_annotated$ENSEMBL[1:N]
sampleInfo <- as.data.frame(colData(dds)[,c("Status","CellType")])

vsd <- vst(dds)
pheatmap(assay(vsd)[top_genes,],
         annotation_col = sampleInfo,scale="row")
```

# Exercise

Modifying the plot for `ENSMUSG00000000381` so that variability in `CellType` and `Status` can be seen

This the basic plot:-

```{r}
dds$Status <- as.factor(dds$Status)
plotCounts(dds, "ENSMUSG00000000381",intgroup = "Status")
```

`plotCounts` allows us to specify multiple groups in the `intgroup` parameter.

```{r}
plotCounts(dds, "ENSMUSG00000000381",intgroup = c("Status","CellType"),returnData = TRUE) %>% 
  ggplot(aes(x = Status,y=count,col=CellType)) + geom_jitter(width=0.1) + scale_y_log10()

```

The `org.Dr.eg.db` package can be used to annotate Zebrafish data

```{r eval=FALSE}
BiocManager::install("org.Dr.eg.db")
```

```{r}
library(biomaRt)
listMarts()
ensembl=useMart("ENSEMBL_MART_ENSEMBL")
listDatasets(ensembl) %>% filter(grepl("Zebra",description))
```

```{r}
ensembl = useDataset("drerio_gene_ensembl", mart=ensembl)
```

