---
title: "R Notebook"
output: html_notebook
---


# Importing the data

Read the raw counts using code from previous session

```{r}
library(tximport)
library(DESeq2)
dirs <- list.files(file.path("salmon_quant/"))
quant_files <- paste0("salmon_quant/",dirs,"/quant.sf.gz")
names(quant_files) <- dirs

tx2gene <- read.csv("tx2gene.csv")

txi <- tximport(quant_files,type="salmon",tx2gene = tx2gene,ignoreTxVersion = TRUE)

sampleinfo <- read.delim("meta_data/sampleInfo_corrected.txt")
rownames(sampleinfo) <- sampleinfo$run

dds <- DESeqDataSetFromTximport(txi, 
                                colData = sampleinfo,
                                design = ~condition)

```



# Exercise

We now want to use the `Status` as our factor of interest. This can be achieved by changing the `design` our object

```{r}
dds_condition<- dds
colData(dds_condition)
## You will need to change this line to choose the correct comparison
design(dds_condition) <- ~condition
```

The default contrast is already TGF vs CTRL. As above we can create a data frame for further processing.

```{r}
de_condition <- DESeq(dds_condition)
results_condition <- results(de_condition) %>%
  as.data.frame %>% 
  tibble::rownames_to_column("ENSEMBL")  

results_condition
saveRDS(results_condition,"Robjects/results_condition.rds")
```

We can filter to discover differentially-expressed genes and write out to a file

```{r}
library(readr)
results_condition %>% filter(abs(log2FoldChange) > 1, padj < 0.05) %>% 
  write_csv("TGF_vs_CTR_DE.csv")
```

# Exercise

In the same way as the workshop, we can obtain annotation using the `org.Hs.eg.db` package

```{r}
library(org.Hs.eg.db)
anno <- AnnotationDbi::select(org.Hs.eg.db,keys=results_condition$ENSEMBL,
            columns=c("ENSEMBL","SYMBOL","GENENAME","ENTREZID"),
            keytype="ENSEMBL") %>% 
filter(!duplicated(ENSEMBL))

results_tgf <- left_join(results_condition,anno)

## save as an R object
saveRDS(results_tgf, file="Robjects/results_TGF_vs_CTR_annotated.rds")
```

We can filter the table to our genes of interest. All seem to have significant p-values in the comparison (as expected).


```{r}
filter(results_tgf, SYMBOL %in% c("COL1A1","COL1A2","COL3A1"))
```

The counts can be visualised using the `plotCounts` function. The `mfrow` function is the "base" graphics method for combining plots - in this case putting three plots in a row.

```{r}
par(mfrow=c(1,3))
plotCounts(dds, "ENSG00000108821", intgroup = "condition")
plotCounts(dds, "ENSG00000164692", intgroup = "condition")
plotCounts(dds, "ENSG00000168542", intgroup = "condition")
```

A ggplot2 version of the plot is a bit more complicated. We first have to extract the relevant normalised gene intensites from the variance-stabilised object. This matrix has *ENSEMBL* IDs for rows.

```{r}
my_genes <- c("ENSG00000108821","ENSG00000164692","ENSG00000168542")
vsd <- vst(dds)
assay(vsd) %>% data.frame %>% 
  tibble::rownames_to_column("ENSEMBL") %>% 
  filter(ENSEMBL %in% my_genes)
```

The matrix is not in the correct shape for `ggplot2`; as it is *wide* rather than *long*. So we will have to use `tidyr` to reshape.

```{r}
assay(vsd) %>% data.frame %>% 
  tibble::rownames_to_column("ENSEMBL") %>% 
  filter(ENSEMBL %in% my_genes) %>% 
  tidyr::pivot_longer(-ENSEMBL, names_to = "Sample",values_to = "count")

```

R isn't keen on having column names that start with a number, so has decided to be an `X` in front of every sample name. We can remove this using `gsub` and then combine with the sample information

```{r}
assay(vsd) %>% data.frame %>% 
  tibble::rownames_to_column("ENSEMBL") %>% 
  filter(ENSEMBL %in% my_genes) %>% 
  tidyr::pivot_longer(-ENSEMBL, names_to = "Run",values_to = "count") %>% 
  mutate(Run  = gsub("X","", Run)) %>% 
  left_join(sampleinfo)
```

Finally, we can plot using a jitter plot and *facet*ing by the gene name.

```{r}
assay(vsd) %>% data.frame %>% 
  tibble::rownames_to_column("ENSEMBL") %>% 
  filter(ENSEMBL %in% my_genes) %>% 
  tidyr::pivot_longer(-ENSEMBL, names_to = "Run",values_to = "count") %>% 
  mutate(Run  = gsub("X","", Run)) %>% 
  left_join(sampleinfo) %>% 
  ggplot(aes(x = condition, y = count, col=condition)) + geom_jitter(width=0.1) + facet_wrap(~ENSEMBL)

```

If we wanted the gene `SYMBOL` instead of `ENSEMBL`, we could incorporate our `anno` data frame before plotting. This will add the `SYMBOL` for these genes of interest.

```{r}
assay(vsd) %>% data.frame %>% 
  tibble::rownames_to_column("ENSEMBL") %>% 
  filter(ENSEMBL %in% my_genes) %>% 
  tidyr::pivot_longer(-ENSEMBL, names_to = "Run",values_to = "count") %>% 
  mutate(Run  = gsub("X","", Run)) %>% 
  left_join(sampleinfo) %>% 
  left_join(anno) %>% 
  ggplot(aes(x = condition, y = count, col=condition)) + geom_jitter(width=0.1) + facet_wrap(~SYMBOL)
```

