---
title: "R Notebook"
output: html_notebook
---


# Importing the data

Read the raw counts using code from previous session

```{r}
library(readr)
library(tximport)
library(DESeq2)
dirs <- list.files(file.path("salmon_quant/"))
quant_files <- paste0("salmon_quant/",dirs,"/quant.sf.gz")
names(quant_files) <- dirs

tx2gene <- read_csv("tx2gene.csv",col_names = FALSE)

txi <- tximport(quant_files,type="salmon",tx2gene = tx2gene)

sampleinfo <- read.delim("meta_data/sampleInfo_corrected.txt")


dds <- DESeqDataSetFromTximport(txi, 
                                colData = sampleinfo,
                                design = ~condition)

```



# Exercise

We now want to use the `condition` as our factor of interest. This can be achieved by changing the `design` our object

```{r}
dds_condition<- dds
design(dds_condition) <- ~condition
```
```{r}
results(de_condition)
```

The default contrast is already TGF vs CTRL. As above we can create a data frame for further processing.

```{r}
results(de_condition, tidy = TRUE)
```

We can filter to discover differentially-expressed genes and write out to a file

```{r}
library(readr)
results(de_condition, tidy=TRUE) %>% 
  filter(abs(log2FoldChange) > 1, padj < 0.05) %>% 
  write_csv("de_analysis/TGF_vs_CTR_DE.csv")
```

# Exercise

In the same way as the workshop, we can obtain annotation using the `org.Hs.eg.db` package

```{r}
results_condition <- results(de_condition,tidy = TRUE)
library(org.Hs.eg.db)
anno <- AnnotationDbi::select(org.Hs.eg.db,keys=results_condition$row,
            columns=c("ENSEMBL","SYMBOL","GENENAME","ENTREZID"),
            keytype="ENSEMBL") %>% 
filter(!duplicated(ENSEMBL))

results(de_condition, tidy=TRUE) %>% left_join(anno,by=c("row"="ENSEMBL"))


```

We can filter the table to our genes of interest. All seem to have significant p-values in the comparison (as expected).


```{r}
results(de_condition, tidy=TRUE) %>% left_join(anno,by=c("row"="ENSEMBL")) %>% 
  filter(SYMBOL %in% c("COL1A1","COL1A2","COL3A1"))
```

The counts can be visualised using the `plotCounts` function. The `mfrow` function is the "base" graphics method for combining plots - in this case putting three plots in a row.

```{r}
par(mfrow=c(1,3))
plotCounts(dds, "ENSG00000108821", intgroup = "condition")
plotCounts(dds, "ENSG00000164692", intgroup = "condition")
plotCounts(dds, "ENSG00000168542", intgroup = "condition")
```

A ggplot2 version of the plot is a bit more complicated. Firstly, we can make use of the `returnData` argument to save the counts used in the plots and create three separate data frames.

```{r}
plot1 <- plotCounts(dds, "ENSG00000108821", intgroup = "condition",returnData = TRUE)
plot2 <- plotCounts(dds, "ENSG00000164692", intgroup = "condition", returnData = TRUE)
plot3 <- plotCounts(dds, "ENSG00000168542", intgroup = "condition", returnData = TRUE)
```

However, we will want to make separate plots for each gene and none of the data frames we have created contain information about what gene they relate to.

e.g.

```{r}
plot1
```

So we have to mutate each data frame to include a column with the gene name

```{r}
plot1 <- plotCounts(dds, "ENSG00000108821", intgroup = "condition",returnData = TRUE) %>% 
  mutate(Gene = "COL1A1")
plot2 <- plotCounts(dds, "ENSG00000164692", intgroup = "condition", returnData = TRUE) %>% 
  mutate(Gene = "COL1A2")
plot3 <- plotCounts(dds, "ENSG00000168542", intgroup = "condition", returnData = TRUE) %>% 
  mutate(Gene = "COL3A1")
```

We can then create a combined data frame with the data from all genes

```{r}
bind_rows(plot1, plot2, plot3)
```
The `ggplot` function can be then be used for plotting. A `geom_jitter` geom adds some random noise to the x-axis and the `facet_wrap` function creates a separate panel for each gene.

```{r}
library(ggplot2)
bind_rows(plot1, plot2, plot3) %>% 
  ggplot(aes(x = condition, y = count, col = condition)) + geom_jitter(width=0.1) + facet_wrap(~Gene)
```
Since we are only looking at CTR and TGF groups, we could choose to exclude `IR` samples by filtering prior to plotting

```{r}
library(ggplot2)
bind_rows(plot1, plot2, plot3) %>% 
  filter(condition != "IR") %>% 
  ggplot(aes(x = condition, y = count, col = condition)) + geom_jitter(width=0.1) + facet_wrap(~Gene)
```
